---
title: "Lab 4"
author: "Sylvia Baeyens"
date: "9/17/2021"
output: github_document
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages, echo= FALSE, include= FALSE}
#including necessary libraries
library(data.table)
library(dplyr)
library(tidyverse)
library(leaflet)
library(R.utils)
library(ggplot2)
```

Objective= to examine the association between weekly average dew pt temp & wind speed in 4 regions of the US

# 1. Read in the Data

```{r data-read, cache= TRUE}
#download file ONLY IF it has not been downloaded previously
if (!file.exists("../met_all.gz")) {
  download.file("https://raw.githubusercontent.com/USCbiostats/data-science-data/master/02_met/met_all.gz", "../met_all.gz", method="libcurl", timeout = 60)}
met <- data.table::fread("../met_all.gz")
```

# 2. Prepare the data
```{r data-prep, cache= TRUE}
#Remove temperatures less than -17C
met = met[temp>-17]
#Make sure there are no missing data in the key variables coded as 9999, 999, etc
met[,table(is.na(temp))]
met[,table(is.na(rh))]
met[,table(is.na(wind.sp))]
met[,table(is.na(vis.dist))]
met[,table(is.na(dew.point))]
met[,table(is.na(lat))]
met[,table(is.na(lon))]
met[,table(is.na(elev))]

met[elev == 9999.0, elev := NA]

#Generate a date variable using the functions as.Date() (hint: You will need the following to create a date paste(year, month, day, sep = "-")).
met[, ymd := as.Date(paste(year, month, day, sep = "-"))]
#Using the data.table::week function, keep the observations of the first week of the month.
met[, table(week(ymd))]

met = met[week(ymd) == 31]
#Compute the mean by station of the variables temp, rh, wind.sp, vis.dist, dew.point, lat, lon, and elev.
met_avg = met[,.(
  temp = mean(temp,na.rm = TRUE),
  rh = mean(rh,na.rm = TRUE),
  wind.sp = mean(wind.sp,na.rm = TRUE),
  vis.dist = mean(vis.dist,na.rm = TRUE),
  dew.point = mean(dew.point,na.rm = TRUE),
  lat = mean(lat,na.rm = TRUE),
  lon = mean(lon,na.rm = TRUE),
  elev = mean(elev,na.rm = TRUE)
),by = "USAFID"]
#Create a region variable for NW, SW, NE, SE based on lon = -98.00 and lat = 39.71 degrees
met_avg[lat>= 39.71 & lon<= -98,region:="Northwest"]
met_avg[lat< 39.71 & lon<= -98,region:="Southwest"]
met_avg[lat>= 39.71 & lon> -98,region:="Northeast"]
met_avg[lat< 39.71 & lon> -98,region:="Southeast"]

met_avg[,table(region, useNA="always")]
#Create a categorical variable for elevation as in the lecture slides
met_avg[, elev_cat:=fifelse(elev>252,"high","low")]
```

# 3. Using geom_violin to examine wind speed & dew pt temp
```{r}
ggplot(met_avg, mapping = aes(y= wind.sp, x=1)) +
  geom_violin() +
  facet_grid(~region)

ggplot(met_avg, mapping = aes(y= dew.point, x=1)) +
  geom_violin() +
  facet_grid(~region)
```
The northeast has the greatest range of wind speed values. However, the average wind speed is larger in both western regions.
The average dewpoint temp is greatest in the southeast region. There is a greater range of dewpoint temp values in both western regions. 

# 4. Using geom_jitter with stat_smooth between dew pt & wind speed
```{r}
ggplot(met_avg, aes(y=wind.sp, x=dew.point)) +
  geom_jitter(aes(color=region)) +
  stat_smooth()

```
There is a slight negative relationship between dew point and wind speed, except for at the highest dew point temperatures. This graph also confirms that the South east has the highest dew point temperature values. 

# 5. Using geom_bar
```{r}
met_avg[] %>%
  ggplot() +
    geom_bar(mapping = aes(x=elev_cat, fill=region), position = "dodge") +
    scale_fill_brewer(palette = "Pastel1") +
    ggtitle('Visualization of Weather Stations by Elevation Category and Region') +
    xlab('Elevation Category') +
    ylab('Number of Weather Stations')
```
At high elevation, there are between 150 and 250 weather stations in each region. The most high elevation weather stations exist in the Northeast, followed by the Southwest. At low elevation, there is a more unequal spread. There are about 500 low elevation weather stations in the Southeast and almost no low elevation stations in the Northwest.

# 6. Using stat_summary to examine mean dew point and wind speed
```{r}
ggplot(met_avg[!is.na(wind.sp) & !is.na(dew.point)], 
       mapping=aes(x=region, y= wind.sp)) +
    stat_summary(fun.data = "mean_sdl") +
    stat_summary(fun.data = "mean_sdl", geom= "errorbar")

ggplot(met_avg[!is.na(wind.sp) & !is.na(dew.point)], 
       mapping=aes(x=region, y= dew.point)) +
    stat_summary(fun.data = "mean_sdl") +
    stat_summary(fun.data = "mean_sdl", geom= "errorbar")
```

Dew point temperature is the temperature at which the air cannot hold any more water in gas form. 

Wind speed is the max speed at which the wind flows as measured by the weather stations. In the northwest and the southwest, the average max wind speed is 3 mph (kmph?) whereas it's close to 2 mph in the northeast and southeast. 

# 7. Making a map!
```{r}
temp.pal = colorNumeric(c('darkgreen','goldenrod','brown'), domain = met_avg$rh)


leaflet(met_avg) %>%
  addProviderTiles('CartoDB.Positron') %>%
  addCircles(
    lat = ~lat, lng=~lon,
                                                  # HERE IS OUR PAL!
    label = ~paste0(round(rh,2), ' rh'), color = ~ temp.pal(rh),
    opacity = 1, fillOpacity = 1, radius = 500
    ) %>%
  # And a pretty legend
  addLegend('bottomleft', pal=temp.pal, values=met_avg$rh,
          title='Temperature, C', opacity=1)
```

# 8. Using ggplot extension
```{r}
#downloading package first
devtools::install_github('thomasp85/gganimate')
library(gganimate)


p = ggplot(met, aes(x=temp, y=wind.sp)) +
  geom_point(alpha = 0.7, show.legend = FALSE) +
  # Here comes the gganimate specific bits
  labs(title = 'USAFID: {frame_time}', x = 'Temperature', y = 'Wind Speed') +
  transition_time(USAFID) +
  ease_aes('linear')

library(gifski)
animate(p, renderer = gifski_renderer())

```
Gif above frame speed is irregular! not entirely sure how to fix that!
